---
title: "Polymorphic inversions underlie the shared genetic susceptibility to prevalent common diseases"
author: "Juan R Gonzalez, Carlos Ruiz-Arenas, Alejandro Caceres, Ignasi Moran, Marcos Lopez, Lorena Alonso, Ignacio Tolosana, Marta Guindo-Martinez, Josep M Mercader, Tonu Esko, David Torrents, Josefa Gonzalez, Luis A Perez-Jurado"
output:
  BiocStyle::html_document:
    toc_depth: 3
  BiocStyle::pdf_document:
    toc_depth: 3
---

```{r style, echo=FALSE}
library("knitr")
opts_chunk$set(message = FALSE, error = TRUE, 
               warning = FALSE, cache=TRUE, comment="")
```

# Getting started 

Let us load the required libraries to reproduce paper's results:

```{r load_libraries}
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(nlme)
library(org.Hs.eg.db)
library(clusterProfiler)
library(scoreInvHap)
library(SNPassoc)
library(snpStats)
library(meta)
library(RColorBrewer)
library(OmicCircos)
library(compareGroups)
```

Then, some functions are created to facilitate data analysis and visualizations. The functions are loaded by:

```{r load_functions, cache=FALSE}
source("R/aux_functions.R")
```

The inversion calling is perfomed using `scoreInvHap` Bioconductor package using the imputed data obtained after applying our pipeline described here: https://github.com/isglobal-brge/imputeInversion. The inversion regions of each SNP array or GWAS dataset are first imputed to get all the required SNPs that are used to call inversion genotypes. 

# Geographical inversion distribution in European regions

We first are interested in finding population structure at European level. To this end, the Population Reference Sample study (POPRES) obtained from dbGAP (accesion number phs000145.v4.p2) is used. We select those samples passing QC and having European origin:

```{r load_POPRES}
popres <- read.delim("POPRES/ScoreInvHap_POPRES_final_full.txt",
                     as.is =TRUE, na=c("", NA, " "))
popres <- popres[popres$'STATUS_OVERALL'=="Passed QC",]
```


```{r remove_and_add_new}
popres.old <- popres
ll <- grep("inv", colnames(popres.old))
popres <- popres.old[,-ll]

load("POPRES/scoreInvHapClassDFPOPRES.Rdata")
popres <- merge(popres, scClassDF, by.x="SUBJID",
           by.y="row.names")

popres.eur <- subset(popres,
                     GROUPING_CONTINENTAL_ORIGIN ==
                       "European")
```


This is the number of samples per region

```{r print_regions}
table(popres.eur$"GROUPING_PCA_LABEL2")
table(popres.eur$"EUROPEAN_LABEL")

```

Let's estimate the inversion frequencies per european region

```{r freq_EURreg}
ii <- grep("inv", names(popres.eur))
freq.EURreg <- getFreqs(popres.eur, 
                        index.inv=ii,
                        grouping = "GROUPING_PCA_LABEL2", 
                        sep="")
freq.EURreg
```

```{r plot_EURreg}
xx <- data.frame(freq.EURreg[,-1])
rownames(xx) <- freq.EURreg$pop
pca.EURreg <- prcomp(xx,
                    center = TRUE, scale = TRUE)
p.1000G.pca <- fviz_pca_biplot(pca.EURreg, 
                               repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
p.1000G.pca
```


```{r plot_EURreg2}
xx2 <- xx[-c(1:3), ]
pca.EURreg2 <- prcomp(xx2, center = TRUE, 
                      scale = TRUE)

fviz_pca_biplot(pca.EURreg2, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```


Let us estimate the frequencies North-Center-South

```{r freq_EURreg2}

popres.eur3 <- popres.eur[popres.eur$GROUPING_PCA_LABEL2%in%
                            c("EuropeN", "EuropeNE", "EuropeNW",
                              "Mix",
                              "EuropeS"),]
popres.eur3$pop <- as.factor(popres.eur3$GROUPING_PCA_LABEL2)
levels(popres.eur3$pop) <- c("North", "North",  "North", "South", "Center")

n.europe <- table(popres.eur3$pop)


ii <- grep("inv", names(popres.eur3))
freq.EURreg2 <- getFreqs(popres.eur3, 
                        index.inv=ii,
                        grouping = "pop", 
                        sep="")
freq.EURreg2$pop <- relevel(freq.EURreg2$pop,2)
freq.EURreg2 <- data.frame(freq.EURreg2)
rownames(freq.EURreg2) <- freq.EURreg2$pop
freq.EURreg2
```

## European countries

We select those countries having more than 20 samples to get robust inversion frequencies estimation and remove those countries that are not in Europe.

```{r freq_EURcountry}
tt <- table(popres.eur$COUNTRY_SELF)
sel <- intersect(names(tt)[tt>20],  
                 names(tt)[!names(tt)%in%
                             c("Australia", "Canada",
                               "USA", "CEU")])
popres.eur2 <- subset(popres.eur, popres.eur$COUNTRY_SELF%in%sel &
                        !is.na(popres.eur$COUNTRY_SELF))
``` 


```{r freq_EURcount}
ii <- grep("inv", colnames(popres.eur2))
freq.EURcountry <- getFreqs(popres.eur2,
                            index.inv = ii,
                            grouping = "COUNTRY_SELF",
                            sep="")
freq.EURcountry
```


```{r plot_EURcountry}
xx <- data.frame(freq.EURcountry[,-1])
rownames(xx) <- freq.EURcountry$pop
pca.EURcountry <- prcomp(xx,
                    center = TRUE, scale = TRUE)
fviz_pca_biplot(pca.EURcountry, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```

```{r plotGeographical1}
x <- tidyr::gather(freq.EURreg, 
                   key="inversion", 
                   value="freq", -pop)
colourCount <- length(unique(x$pop)) 
x2 <- x
p2 <- ggplot(x2, aes(x=pop, y=freq, fill=pop)) +
  geom_bar(stat="identity") + facet_wrap(~ inversion) +
  labs(fill = "Region", x="", y="Frequency of I allele") +
  ggtitle("European regions (POPRES)") +
  scale_fill_manual(values = colorRampPalette
                    (brewer.pal(11, "PRGn"))(colourCount))+
  theme(axis.text.x=element_blank(), 
        axis.title.y = element_text(size=16),
        axis.text.y = element_text(size=12),
        title = element_text(face="bold", size=14))  

p2
```

```{r plotGeographical2}
x <- tidyr::gather(freq.EURreg2, 
                   key="inversion", 
                   value="freq", -pop)
x$pop2 <- factor(x$pop, levels=c("North", "Center", "South"))

p3 <- ggplot(x, aes(x=pop2, y=freq, fill=pop2)) +
  geom_bar(stat="identity") + facet_wrap(~ inversion) +
  labs(fill = "Region", x="", y="Frequency of I allele") +
  ggtitle("POPRES") +
  scale_fill_manual(values = colorRampPalette
                    (brewer.pal(11, "PRGn"))(colourCount))+
  theme(axis.text.x=element_blank(), 
        axis.title.y = element_text(size=16),
        axis.text.y = element_text(size=12),
        title = element_text(face="bold", size=14))  

p3
```

```{r test_trend}

ll <- c("North", "Center", "South")
mat <- matrix(NA, nrow=length(invs.pops), ncol=4)
j <- 1
for (i in invs.pops) {
  freqs.i <- freq.EURreg2[ll,i]
  p.i <- prop.trend.test(round((freqs.i/100)*n.europe,0), n.europe, score=c(1,2,3))$p.value
  mat[j,] <- c(freqs.i, p.i)
  j <- j + 1
}
rownames(mat) <- invs.pops
colnames(mat) <- c("North", "Center", "South", "pvalue")
cbind(round(mat[,1:3],1), pval=round(mat[,4],4))

ii <- grep("^inv", names(popres.eur))
x.s <- setupSNP(popres.eur, ii, sep="")
summary(x.s)
```



### European countries

```{r plotEuropeanFreqs}
x <- tidyr::gather(freq.EURcountry, 
                   key="inversion", 
                   value="freq", -pop)
colourCount <- length(unique(x$pop)) 

p1 <- ggplot(x[!x$pop%in%c("EuropeanA", "EuropeanB"),], 
             aes(x=pop, y=freq, fill=pop)) +
      geom_bar(stat="identity") + facet_wrap(~ inversion) +
      labs(fill = "Country", x="", 
           y="Frequency of I allele") +
      ggtitle("POPRES - European Countries") +
  scale_fill_manual(values = colorRampPalette
                    (brewer.pal(11, "PRGn"))(colourCount))+
  theme(axis.text.x=element_blank(), 
        axis.title.y = element_text(size=16),
        axis.text.y = element_text(size=12),
        title = element_text(face="bold", size=14))  

p1
```




# Inversion association studies

## Discovery analysis: UK Biobank

The object having the results (p-values and sign effect) corresponding to the association analyse is loaded by:

```{r load_pUKB}
load("UKBiobank/pUKB.Rdata")
```

NOTE: genotypes and fenotypes to get the p-values are avialable in the [UK Biobank repository](https://biobank.ndph.ox.ac.uk/showcase/docs.cgi?id=1)

Then the Figure 2 in the manuscript is created by

```{r plotUKB, fig.height=8}

o <- pUKBend$p.value < 0.05/length(invs) &
  !is.na(pUKBend$p.value)

pUKBend$p.value[!o] <- NA
pUKBend$sign[!o] <- NA

pltUKB <- ggplot(pUKBend, aes(x = band, y = label, 
                        size=-log10(p.value))) + 
    geom_point(aes(col = sign)) +
    scale_color_manual(values=c("red", "blue"),
                       na.translate=FALSE) +
    xlab("") + ylab("") +
    labs(color="Effect I allele") +
    theme(axis.text.x = element_text(angle = 90, vjust=0.5,
                                     hjust=1)) +
    ggtitle("Association analysis UK Biobank")
pltUKB
```

The Odds Ratios are here:

```{r odds_ukb}
load("UKBiobank/resUKB_final.Rdata")
invs.sel <- c("inv8_001", "inv16_009")
phenos <- c("obesity", "diabetes", "hypertension", "asthma", "depression",
                     "diabetes/obesity", "hypertension/obesity", "asthma/obesity",
                     "depression/obesity")

invOR <- list()
for (i in 1:2){
  a1 <- SNPassoc::odds(ans.ukb$obesity, model="log-add")[invs.sel[i],]
  a2 <- SNPassoc::odds(ans.ukb$diabetes, model="log-add")[invs.sel[i],]
  a3 <- SNPassoc::odds(ans.ukb$hypertension, model="log-add")[invs.sel[i],]
  a4 <- SNPassoc::odds(ans.ukb$asthma, model="log-add")[invs.sel[i],]
  a5 <- SNPassoc::odds(ans.ukb$everDep, model="log-add")[invs.sel[i],]
  a6 <- SNPassoc::odds(ans.ukb$diab.ob, model="log-add")[invs.sel[i],]
  a7 <- SNPassoc::odds(ans.ukb$hyp.ob, model="log-add")[invs.sel[i],]
  a8 <- SNPassoc::odds(ans.ukb$asthma.ob, model="log-add")[invs.sel[i],]
  a9 <- SNPassoc::odds(ans.ukb$ob.dep, model="log-add")[invs.sel[i],]
  invOR[[i]] <- rbind(a1, a2, a3, a4, a5, a6, a7, a8, a9)
  names(invOR[[i]])[4] <- "p.value"
  rownames(invOR[[i]]) <- phenos
}
names(invOR) <- invs.sel
invOR
```



## Validation: GERA

```{r gera}
# NOTE: remove manually first row original file from BSC
load("diabetes/70kT2D.Rdata")
gera.pheno <- read.table("GERA/GERA_BMI.sample", header=TRUE)
gera <- merge(gera, gera.pheno, by.x="SID", by.y="ID_1")
gera.cov <- read.table("GERA/phs000674.v1.pht003641.v1.p1.c1.Survey.HROO.txt", header=TRUE)
gera2 <- merge(gera, gera.cov, by.x="SID", by.y="SUBJID")

gera$inv16_009 <- dplyr::recode(gera$inv16_009, "NN"="II", "II"="NN")

gera$diab.ob <- rep(NA, nrow(gera))
gera$diab.ob[gera$DIA2==0 & gera$ob==0] <- 0
gera$diab.ob[gera$DIA2==1 & gera$ob==1] <- 1

gera$hyper.ob <- rep(NA, nrow(gera))
gera$hyper.ob[gera$HYPER==0 & gera$ob==0] <- 0
gera$hyper.ob[gera$HYPER==1 & gera$ob==1] <- 1

gera$dep.ob <- rep(NA, nrow(gera))
gera$dep.ob[gera$DEPRESS==0 & gera$ob==0] <- 0
gera$dep.ob[gera$DEPRESS==1 & gera$ob==1] <- 1

gera$asthma.ob <- rep(NA, nrow(gera))
gera$asthma.ob[gera$ASTHMA==0 & gera$ob==0] <- 0
gera$asthma.ob[gera$ASTHMA==1 & gera$ob==1] <- 1

ss <- grep("inv", names(gera))
gera.s <- setupSNP(gera, colSNPs = ss, sep="")

obesity.gera <- WGassociation(ob ~ sex + + BIRTHYEARCAT + PC1 +
                              PC2 + PC3 + PC4, data=gera.s, geno=0)


hyper.gera <- WGassociation(HYPER ~ sex + + BIRTHYEARCAT + PC1 +
                              PC2 + PC3 + PC4, data=gera.s, geno=0)
depress.gera <- WGassociation(DEPRESS ~ sex + BIRTHYEARCAT + PC1 +
                                PC2 + PC3 + PC4, data=gera.s)
asthma.gera <- WGassociation(ASTHMA ~ sex + BIRTHYEARCAT + PC1 +
                                PC2 + PC3 + PC4, data=gera.s)

hyper.ob.gera <- WGassociation(hyper.ob ~ sex + BIRTHYEARCAT + PC1 +
                               PC2 + PC3 +  PC4, data=gera.s, geno=0)
dep.ob.gera <- WGassociation(dep.ob ~ sex + BIRTHYEARCAT + PC1 +
                               PC2 + PC3 +  PC4, data=gera.s,
                             geno=0)
ashtma.ob.gera <- WGassociation(asthma.ob ~ sex + BIRTHYEARCAT + PC1 +
                               PC2 + PC3 +  PC4, data=gera.s,
                             geno=0)
diab.ob.gera <- WGassociation(diab.ob ~ sex + BIRTHYEARCAT + PC1 +
                               PC2 + PC3 +  PC4, data=gera.s,
                             geno=0)
```

These are the effects:

```{r odds_gera}
invs.sel <- c("inv8_001", "inv16_009")
SNPassoc::odds(hyper.gera, model="log-add")[invs.sel[i],]
SNPassoc::odds(asthma.gera, model="log-add")[invs.sel[i],]
SNPassoc::odds(depress.gera, model="log-add")[invs.sel[i],]

SNPassoc::odds(hyper.ob.gera, model="log-add")[invs.sel[i],]
SNPassoc::odds(ashtma.ob.gera, model="log-add")[invs.sel[i],]
SNPassoc::odds(dep.ob.gera, model="log-add")[invs.sel[i],]

```



```{r pGERA}
pvalGERA <- NULL
temp <- getP(obesity.gera)
pvalGERA <- c(pvalGERA, temp)

temp <- getP(hyper.gera)
pvalGERA <- c(pvalGERA, temp)

temp <- getP(diabetes.res)
pvalGERA <- c(pvalGERA, temp)


temp <- getP(asthma.gera)
pvalGERA <- c(pvalGERA, temp)


temp <- getP(depress.gera)
pvalGERA <- c(pvalGERA, temp)
```

```{r plot_inflated, fig.height=6, fig.width=6}
par(mfrow=c(1,2), mar=c(4,4.2,2,1.5))
p1 <- pUKB$p.value
qqplotBand(p1[!is.na(p1)], pch=16, main="UKB")
p2 <- pvalGERA
qqplotBand(p2[!is.na(p2)], pch=16, main="GERA + 70KT2D")
```

# Transcriptomic effects in GTeX
